---
title: 'Task — Biochemistry & Oncology: Mutational Impact on Protein Structure & Function'
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options applied to all code chunks
knitr::opts_chunk$set(
  echo    = TRUE,    # display code in rendered output
  message = FALSE,   # suppress package loading messages
  warning = FALSE,   # suppress warnings
  fig.width  = 10,
  fig.height = 6
)
```

---

# 1. Load Packages

```{r packages}
# pacman installs missing packages automatically before loading
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,   # dplyr (data wrangling) + ggplot2 (visualization) + stringr
  ggplot2,     # explicit for clarity
  scales,      # axis label formatting (percent, comma)
  patchwork    # combine multiple plots into panels
)
```

---

# 2. Import the Datasets

```{r import-data}
# ── SIFT dataset ─────────────────────────────────────────────
# SIFT (Sorting Intolerant From Tolerant) predicts whether an amino
# acid substitution affects protein FUNCTION.
# Score < 0.05 → mutation is DELETERIOUS to function

sift <- read.table(
  "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/datasets/sift.tsv",
  header = TRUE,
  sep = "",               # handles both tabs and multiple spaces
  stringsAsFactors = FALSE
)

# ── FoldX dataset ────────────────────────────────────────────
# FoldX estimates the change in protein folding free energy (ΔΔG).
# Score > 2 kCal/mol → mutation is DELETERIOUS to structure

foldx <- read.delim(
  "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/datasets/foldX.tsv",
  header = TRUE,
  sep = "", 
  stringsAsFactors = FALSE
)

# ── Quick overview ───────────────────────────────────────────
cat("=== SIFT Dataset ===\n")
dim(sift)
head(sift)

cat("\n=== FoldX Dataset ===\n")
dim(foldx)
head(foldx)
```

> **What we observe:**
> Both datasets share two key columns: `Protein` (UniProt accession ID)
> and `Amino_Acid` (mutation in standard nomenclature e.g. `E63D`).
> The SIFT dataset has scores from 0 to 1; FoldX has scores that can be
> negative (stabilising) or positive (destabilising).

---

# 3. Create the `specific_Protein_aa` Column

```{r specific-protein-aa}
# ── Concatenate Protein + Amino_Acid with underscore separator ──
# This creates a unique identifier for each mutation per protein
# Example: Protein = "A5A607", Amino_Acid = "E63D"
#          → specific_Protein_aa = "A5A607_E63D"

sift <- sift %>%
  mutate(specific_Protein_aa = paste(Protein, Amino_Acid, sep = "_"))

foldx <- foldx %>%
  mutate(specific_Protein_aa = paste(Protein, Amino_Acid, sep = "_"))

# ── Verify the new column ────────────────────────────────────
cat("=== SIFT — new column preview ===\n")
head(sift$specific_Protein_aa, 10)

cat("\n=== FoldX — new column preview ===\n")
head(foldx$specific_Protein_aa, 10)
```

> The `specific_Protein_aa` column now acts as a **unique key** that
> identifies exactly which amino acid position in which protein was mutated.
> This is critical for correctly merging the two datasets — without it,
> the same mutation code (e.g. `E63D`) could match across different proteins.

---

# 4. Merge SIFT and FoldX Datasets

```{r merge-datasets}
# ── Inner join on the specific_Protein_aa key ────────────────
# Inner join keeps ONLY mutations present in BOTH datasets.
# This ensures every row has both a SIFT score and a FoldX score.

merged_df <- inner_join(
  sift  %>% 
    select(specific_Protein_aa, Protein, Amino_Acid, sift_Score),
  foldx %>%
    select(specific_Protein_aa, foldX_Score),
  by = "specific_Protein_aa"
)

write.csv(merged_df, "Dataset_cleaned.csv", row.names = FALSE)

# ── Check the merged result ──────────────────────────────────
cat("=== Merged Dataset ===\n")
dim(merged_df)
head(merged_df, 10)
colSums(is.na(merged_df))   # check for any missing values after merge
```

> **After merging:**
> Only mutations that have **both** a SIFT score and a FoldX score are
> retained. Any mutation measured in only one dataset is excluded.
> The final dataframe has 5 meaningful columns:
> `specific_Protein_aa`, `Protein`, `Amino_Acid`, `sift_Score`, `foldX_Score`.

---

# 5. Filter Deleterious Mutations (Both Structure AND Function)

```{r deleterious-filter}
# ── Apply the authors' criteria ──────────────────────────────
# SIFT  < 0.05       → mutation disrupts protein FUNCTION
# FoldX > 2 kCal/mol → mutation disrupts protein STRUCTURE
#
# We want mutations that are deleterious in BOTH dimensions simultaneously

deleterious <- merged_df %>%
  filter(sift_Score  < 0.05,
         foldX_Score > 2.00)

write.csv(deleterious, "Deleterious.csv", row.names = FALSE)

# ── How many such mutations exist? ───────────────────────────
cat(sprintf("Total mutations in merged dataset : %d\n", nrow(merged_df)))
cat(sprintf("Doubly deleterious mutations      : %d\n", nrow(deleterious)))
cat(sprintf("Percentage                        : %.2f%%\n",
            nrow(deleterious) / nrow(merged_df) * 100))

# ── Preview the deleterious mutations ───────────────────────
head(deleterious, 20)
```

> **Observation:** Only a **small fraction** of all mutations are doubly
> deleterious — affecting both protein structure and function simultaneously.
> This aligns with the biological principle that most mutations are neutral
> or only mildly disruptive. The ones identified here are the most
> biologically dangerous variants — candidates for disease-causing mutations.

---

# 6. Amino Acid Substitution Nomenclature

```{r aa-nomenclature}
# ── Understanding the Amino_Acid column format ───────────────
# Standard HGVS-like protein substitution notation: XposY
#   X   = original (wild-type) amino acid (single-letter code)
#   pos = position in the protein sequence (integer)
#   Y   = new (mutant) amino acid (single-letter code)
#
# Example: "E63D"
#   E = Glutamic acid (wild-type) at position 63
#   D = Aspartic acid (mutant)
#
# To find the amino acid with the MOST structural + functional impact,
# we extract the FIRST CHARACTER of Amino_Acid → the original AA

# ── Extract the wild-type (original) amino acid ──────────────
# substr(string, start, stop) extracts characters by position
# substr(x, 1, 1) → first character only

deleterious <- deleterious %>%
  mutate(wt_aa = substr(Amino_Acid, 1, 1))

# ── Preview ──────────────────────────────────────────────────
head(deleterious %>% select(Amino_Acid, wt_aa), 15)

# ── Map single-letter codes to full amino acid names ─────────
aa_names <- c(
  A = "Alanine",       C = "Cysteine",    D = "Aspartic Acid",
  E = "Glutamic Acid", F = "Phenylalanine",G = "Glycine",
  H = "Histidine",     I = "Isoleucine",  K = "Lysine",
  L = "Leucine",       M = "Methionine",  N = "Asparagine",
  P = "Proline",       Q = "Glutamine",   R = "Arginine",
  S = "Serine",        T = "Threonine",   V = "Valine",
  W = "Tryptophan",    Y = "Tyrosine"
)

deleterious <- deleterious %>%
  mutate(wt_aa_name = aa_names[wt_aa])
```

> **Nomenclature note:**
> The first character of the `Amino_Acid` code represents the **original
> amino acid** before mutation. By extracting and counting these, we can
> identify which wild-type amino acids — when mutated — cause the most
> simultaneous structural and functional damage.

---

# 7. Frequency Table of Amino Acids

```{r freq-table}
# ── Count how many doubly deleterious mutations each WT AA has ─
aa_freq <- deleterious %>%
  count(wt_aa, wt_aa_name, name = "Frequency") %>%
  arrange(desc(Frequency))

# ── Print the frequency table ────────────────────────────────
print(aa_freq)

# ── Identify the amino acid with the HIGHEST impact ──────────
top_aa <- aa_freq %>% slice(1)
cat(sprintf(
  "\nAmino acid with the most structural & functional impact:\n→ %s (%s) with %d deleterious mutations\n",
  top_aa$wt_aa, top_aa$wt_aa_name, top_aa$Frequency
))

# ── Amino acids with more than 100 occurrences ───────────────
high_impact <- aa_freq %>% filter(Frequency > 100)
cat("\nAmino acids with more than 100 deleterious mutations:\n")
print(high_impact)
```

> **Observation:**
> The frequency table tells us which original amino acids, when substituted,
> most frequently produce mutations that are deleterious to BOTH structure
> and function. High-frequency amino acids tend to occupy **critical positions**
> in the protein — often buried hydrophobic cores or active sites where
> any change disrupts folding or catalytic activity.

---

# 8. Bar Plot — Amino Acid Frequency

```{r barplot}
# ── Define a clean consistent theme ─────────────────────────
theme_bio <- theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, face = "bold")
  )

# ── Bar plot ─────────────────────────────────────────────────
barplot_aa <- ggplot(aa_freq,
                     aes(x    = reorder(wt_aa_name, -Frequency),
                         y    = Frequency,
                         fill = Frequency)) +
  geom_col(color = "white", width = 0.75) +
  geom_text(aes(label = Frequency),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  # Highlight amino acids with > 100 occurrences
  geom_hline(yintercept = 100, linetype = "dashed",
             color = "red", linewidth = 0.8) +
  annotate("text", x = 1, y = 105,
           label = "Threshold: 100 mutations",
           color = "red", size = 3.5, hjust = 0) +
  scale_fill_gradient(low = "#90CAF9", high = "#1A237E") +
  labs(
    title    = "Frequency of Wild-Type Amino Acids in Doubly Deleterious Mutations",
    subtitle = "Mutations with SIFT < 0.05 AND FoldX > 2 kCal/mol",
    x        = "Wild-Type Amino Acid",
    y        = "Number of Deleterious Mutations",
    fill     = "Frequency",
    caption  = "Source: Galardini et al. | HackBio Dataset"
  ) +
  theme_bio

print(barplot_aa)

ggsave("bar_plot.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: histograms.png\n")
```

> **Observation from bar plot:**
> The bar plot reveals a clear hierarchy — some amino acids appear in
> deleterious mutations far more frequently than others. Amino acids
> with tall bars are those whose substitution is most likely to
> simultaneously disrupt protein stability and function.
> The red dashed line marks the **100-mutation threshold** to visually
> separate high-impact from moderate-impact amino acids.

---

# 9. Pie Chart — Amino Acid Frequency

```{r piechart}
# ── Compute percentage for each amino acid ───────────────────
aa_freq_pct <- aa_freq %>%
  mutate(
    Percentage = round(Frequency / sum(Frequency) * 100, 1),
    Label      = paste0(wt_aa, "\n", Percentage, "%")
  )

# ── Pie chart ────────────────────────────────────────────────
pie_aa <- ggplot(aa_freq_pct,
                 aes(x    = "",
                     y    = Frequency,
                     fill = wt_aa_name)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y") +                # convert bar → pie
  geom_text(aes(label = ifelse(Percentage >= 3, Label, "")),
            position = position_stack(vjust = 0.5),
            size = 3.2, fontface = "bold", color = "white") +
  scale_fill_manual(values = colorRampPalette(
    c("#1A237E", "#1565C0", "#1E88E5", "#42A5F5",
      "#90CAF9", "#E3F2FD", "#B71C1C", "#E53935",
      "#EF9A9A", "#1B5E20", "#2E7D32", "#66BB6A",
      "#F57F17", "#FFA000", "#FFD54F", "#4A148C",
      "#7B1FA2", "#CE93D8", "#880E4F", "#E91E63")
  )(nrow(aa_freq_pct))) +
  labs(
    title    = "Proportion of Wild-Type Amino Acids in Doubly Deleterious Mutations",
    subtitle = "Labels shown for amino acids ≥ 3% of total",
    fill     = "Amino Acid",
    caption  = "Source: Galardini et al. | HackBio Dataset"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(color = "grey50", size = 11, hjust = 0.5),
    legend.position = "right"
  )

print(pie_aa)

ggsave("Pie.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: piee diagram.png\n")
```

---

# 10. Combined Panel Plot

```{r combined-panel, fig.height=8}
# ── Display bar and pie side by side using patchwork ─────────
combined_plot <- barplot_aa + pie_aa +
  plot_annotation(
    title   = "Impact of Wild-Type Amino Acids on Protein Structure & Function",
    caption = "Doubly deleterious: SIFT < 0.05 AND FoldX > 2 kCal/mol",
    theme   = theme(plot.title = element_text(face = "bold", size = 16))
  )

print(combined_plot)

ggsave("graph_combined.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: Combinaison of both graphs.png\n")
```

---

# 11. Description — Amino Acid with the Highest Impact

```{r top-aa-description}
# Print a biological description of the top amino acid
cat("=== TOP AMINO ACID — BIOLOGICAL DESCRIPTION ===\n\n")

cat(sprintf("Amino Acid : %s (%s)\n", top_aa$wt_aa, top_aa$wt_aa_name))
cat(sprintf("Frequency  : %d doubly deleterious mutations\n\n", top_aa$Frequency))

# Descriptions for the most commonly observed top amino acids
aa_descriptions <- list(
  R = paste(
    "ARGININE (R) is a positively charged, hydrophilic amino acid with a",
    "large guanidinium side chain. It is critical for:",
    "  • Salt bridges and electrostatic interactions that stabilise protein 3D structure",
    "  • DNA/RNA binding interfaces due to its positive charge",
    "  • Enzyme active sites where it coordinates substrates or phosphate groups",
    "  • Signal transduction — many kinase substrates contain Arg residues",
    "When Arginine is mutated, both the structural electrostatic network AND",
    "the functional binding capacity of the protein are disrupted simultaneously.",
    sep = "\n"
  ),
  S = paste(
    "SERINE (S) is a small, polar, hydroxyl-containing amino acid. It is critical for:",
    "  • Phosphorylation sites — Serine is the most commonly phosphorylated residue",
    "  • Hydrogen bonding networks that maintain secondary structure",
    "  • Active sites of serine proteases (catalytic triad: Ser-His-Asp)",
    "  • Post-translational modification hubs (glycosylation, acetylation)",
    "Serine mutations disrupt both the structural hydrogen bond network",
    "and critical enzymatic or signalling functions.",
    sep = "\n"
  ),
  G = paste(
    "GLYCINE (G) is the smallest amino acid with no side chain (H), giving it",
    "exceptional conformational flexibility. It is critical for:",
    "  • Tight turns and loops in protein backbone — only AA that fits in restricted angles",
    "  • Collagen triple helix (Gly-X-Y repeat requires every 3rd residue to be Gly)",
    "  • Hinge regions enabling domain movement in enzymes",
    "Glycine mutations force a bulkier side chain into constrained backbone positions,",
    "severely disrupting both the 3D fold and the protein's dynamic function.",
    sep = "\n"
  ),
  P = paste(
    "PROLINE (P) is a unique cyclic amino acid that introduces rigidity and kinks",
    "in polypeptide chains. It is critical for:",
    "  • Structural turns and bends — creates cis peptide bonds unlike other AAs",
    "  • Collagen structure (every 3rd position in Gly-Pro-Hyp repeats)",
    "  • Protein-protein interaction interfaces",
    "Proline mutations remove the structural kink, unfolding loops and",
    "disrupting binding surfaces simultaneously.",
    sep = "\n"
  )
)

# Print the description for the top amino acid, or a generic message
if (top_aa$wt_aa %in% names(aa_descriptions)) {
  cat(aa_descriptions[[top_aa$wt_aa]])
} else {
  cat(sprintf(
    "%s (%s) appears frequently in doubly deleterious mutations, suggesting\n",
    top_aa$wt_aa, top_aa$wt_aa_name
  ))
  cat("it occupies structurally or functionally critical positions in many proteins.\n")
}
```

---

# 12. Amino Acids with More Than 100 Occurrences

```{r high-impact-analysis}
cat("=== AMINO ACIDS WITH > 100 DELETERIOUS MUTATIONS ===\n\n")
print(high_impact)

cat("\n--- STRUCTURAL PROPERTY ---\n")
cat(paste(
  "Amino acids appearing in more than 100 doubly deleterious mutations tend to be",
  "structurally CRITICAL residues. They often:",
  "  • Occupy buried positions in the protein hydrophobic core",
  "  • Form key hydrogen bonds or salt bridges that maintain tertiary structure",
  "  • Sit at domain-domain interfaces where any change destabilises the fold",
  "  • Have restricted conformational space — any substitution strains the backbone",
  "Their high FoldX scores (> 2 kCal/mol) confirm that substituting them",
  "significantly INCREASES the free energy of folding → the protein becomes",
  "thermodynamically LESS STABLE and more prone to misfolding or aggregation.",
  sep = "\n"
))

cat("\n\n--- FUNCTIONAL PROPERTY ---\n")
cat(paste(
  "Functionally, these high-frequency amino acids are:",
  "  • Located in or near active sites, binding pockets, or allosteric regions",
  "  • Involved in substrate recognition, catalysis, or cofactor binding",
  "  • Conserved across species (low SIFT scores confirm evolutionary intolerance)",
  "Their low SIFT scores (< 0.05) confirm that these positions are under",
  "STRONG PURIFYING SELECTION — nature rarely tolerates changes here because",
  "any substitution abolishes or severely impairs the protein's biological function.",
  sep = "\n"
))

cat("\n\n--- COMBINED CONCLUSION ---\n")
cat(paste(
  "Amino acids with > 100 occurrences in doubly deleterious mutations represent",
  "the most VULNERABLE positions in the human proteome — residues where a single",
  "point mutation can simultaneously unfold the protein AND destroy its function.",
  "These are prime candidates for:",
  "  • Disease-causing variants in human genetic disorders",
  "  • Drug target design (stabilising these positions therapeutically)",
  "  • Protein engineering (avoiding these positions in synthetic proteins)",
  sep = "\n"
))
```

---

# 13. Visualise Scores Distribution

```{r score-distributions, fig.height=5}
# ── Scatter plot: SIFT vs FoldX for all merged mutations ─────
# Colour-code by mutation category
merged_df <- merged_df %>%
  mutate(Category = case_when(
    sift_Score  < 0.05 & foldX_Score > 2 ~ "Both Deleterious",
    sift_Score  < 0.05 & foldX_Score <= 2 ~ "Function Only",
    sift_Score >= 0.05 & foldX_Score > 2  ~ "Structure Only",
    TRUE                                   ~ "Neutral"
  ))

scatter_scores <- ggplot(merged_df,
                         aes(x     = sift_Score,
                             y     = foldX_Score,
                             color = Category)) +
  geom_point(alpha = 0.4, size = 1.2) +
  # Threshold lines
  geom_vline(xintercept = 0.05, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  geom_hline(yintercept = 2.00, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  # Annotate quadrants
  annotate("text", x = 0.01, y = max(merged_df$foldX_Score) * 0.95,
           label = "Both\nDeleterious", color = "#B71C1C",
           fontface = "bold", size = 3.5) +
  scale_color_manual(values = c(
    "Both Deleterious" = "#B71C1C",
    "Function Only"    = "#F57F17",
    "Structure Only"   = "#1565C0",
    "Neutral"          = "grey70"
  )) +
  labs(
    title    = "SIFT Score vs FoldX Score for All Mutations",
    subtitle = "Dashed lines = thresholds (SIFT < 0.05 | FoldX > 2)",
    x        = "SIFT Score (< 0.05 = deleterious to function)",
    y        = "FoldX Score in kCal/mol (> 2 = deleterious to structure)",
    color    = "Mutation Category",
    caption  = "Source: Galardini et al. | HackBio Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(scatter_scores)

ggsave("Scatter_scores.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: Scatter_scores.png\n")
```

---

# 14. Overall Conclusions

> **1. Merge strategy:** Using `specific_Protein_aa` as a composite key
> ensures mutations are correctly matched — the same mutation code can
> exist in different proteins, so concatenating Protein + Amino_Acid
> prevents false matches.

> **2. Doubly deleterious mutations are rare:** Only a small percentage
> of all mutations satisfy both SIFT < 0.05 AND FoldX > 2, confirming
> that simultaneous structural and functional disruption is the exception,
> not the rule.

> **3. Top amino acid:** The amino acid with the highest impact (highest
> frequency in doubly deleterious mutations) tends to be one that occupies
> **evolutionarily conserved, structurally critical positions** — often
> involved in electrostatic networks, active sites, or backbone geometry.

> **4. Amino acids with > 100 occurrences** combine two properties:
> they are **structurally essential** (FoldX confirms destabilisation)
> and **functionally irreplaceable** (SIFT confirms evolutionary intolerance).
> These represent the most disease-relevant mutation hotspots in the proteome.

> **5. Clinical relevance:** Identifying these amino acids supports
> precision medicine — variants at these positions in patient genomes
> are strong candidates for pathogenic mutations worth clinical investigation.

---

*End of Task — Biochemistry & Oncology Mutational Impact Analysis*
